---
title: "Cytokinedata_pre_processing and MOFA"
author: "Anna Österberg"
date: "2024-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = TRUE,
                      warning = FALSE)
```

```{r}
#Loading needed libraries 
library(factoextra)
library(FactoMineR)
library(ggplot2)
library(tidyverse)
library(MOFA2)
library(data.table)
library(reticulate)
library(readxl)
```

### Plotting PCA results of PEV samples before and after normalization 

```{r}
#Plotting PCA colored by activation before normalization
raw_data <- read.csv("/home/aosterberg/Documents/DI_työ/cytokine24h.csv",header = TRUE)
colnames(raw_data) <- c("secretome", "Unstimulated_PEV_1", "Unstimulated_PEV_2", "Unstimulated_PEV_3", "Unstimulated_PEV_4","GPVI_PEV_1", "GPVI_PEV_2", "GPVI_PEV_3", "GPVI_PEV_4", "CLEC2_PEV_1","CLEC2_PEV_2","CLEC2_PEV_3", "CLEC2_PEV_4", "TC_PEV_1","TC_PEV_2","TC_PEV_3", "TC_PEV_4")
test_data <- raw_data[,2:17]
cytokine_data <- as.matrix(test_data)

cytokine_groups <- c("Unstimulated", "Unstimulated", "Unstimulated", "Unstimulated", "GPVI", "GPVI","GPVI", "GPVI", "CLEC2", "CLEC2", "CLEC2", "CLEC2","TC","TC","TC","TC")
sample_colors <- c("CLEC2" = "red", "GPVI" = "green", "TC" = "blue", "Unstimulated"= "purple")

pca_data <- PCA(t(cytokine_data), scale.unit = TRUE, graph = FALSE)
fviz_eig(pca_data, addlabels = TRUE)
eigenvalues <- pca_data$eig

fviz_pca_ind(pca_data,
             col.ind = cytokine_groups, # color by groups
             palette = sample_colors, mean.point = FALSE, repel = TRUE,
             legend.title = "Groups")+labs(title = "PCA of PEVs macrophage secretome after 24 h", x= paste0("PC1 (",round(eigenvalues[1,2],2),"%)"), y= paste0("PC2 (",round(eigenvalues[2,2],2),"%)")) + theme_minimal()+ scale_shape_manual(values = rep(16, length(unique(PEVmiRNA_groups)))) + guides(color = guide_legend(override.aes = list(shape = 16, size = 3)))


#We plot boxplots of unscaled and then scaled cytokine values 
cytokine_data[,-1] %>% boxplot
cytokine_data[,-1] %>% scale %>% boxplot


#We create a function to normalize to the standard distribution and apply it to the data
stdnorm <- function(x){
  r = rank(x, ties.method = "random")
  qnorm(r / (length(x) +1))
}

cytokine_norm <- apply(cytokine_data, 2, stdnorm)
head(cytokine_norm)
boxplot(cytokine_norm)


#We then plot the normalized data
pca_data <- PCA(t(cytokine_norm), scale.unit = TRUE, graph = FALSE)
fviz_eig(pca_data, addlabels = TRUE)
eigenvalues <- pca_data$eig

fviz_pca_ind(pca_data,
             col.ind = cytokine_groups, # color by groups
             palette = sample_colors, mean.point = FALSE, repel = TRUE,
             legend.title = "Groups")+labs(title = "PCA of normalized PEVs macrophage secretome after 24 h", x= paste0("PC1 (",round(eigenvalues[1,2],2),"%)"), y= paste0("PC2 (",round(eigenvalues[2,2],2),"%)")) + theme_minimal()+ scale_shape_manual(values = rep(16, length(unique(PEVmiRNA_groups)))) + guides(color = guide_legend(override.aes = list(shape = 16, size = 3)))

```

We notice that unstimulated PEVs and GPVI PEVs cluster on top of each other and does not separate in own clusters as before normalization, therefore we choose not to normalize these values as it removes some biologically relevant information and the intensities does not vary largely between samples. 


## Calculating total intensities 

```{r}

#Loading the cytokine data for both 6h and 24h 
cytokine_6h <- read_excel("/home/aosterberg/Documents/DI_työ/data/TableS2.xlsx", sheet = 2) %>% as.data.frame() %>% column_to_rownames(var="...1")

cytokine_24h <- read_excel("/home/aosterberg/Documents/DI_työ/data/TableS2.xlsx", sheet = 3) %>% as.data.frame() %>% column_to_rownames(var="...1")
  

cytokine_6h <- cytokine_6h[,1:18] #we drop the two last columns with only NA values 

#We caclulate the total intensity and split it into a new dataframe based on PEV group
total_6h <-colSums(cytokine_6h) 
groups_6h <- split(total_6h, rep(1:5, each = 4)) %>% as.data.frame()
groups_6h[3:4,5]<-NA #we add NA values for the missing replicates


total_24h <-colSums(cytokine_24h) 
groups_24h <- split(total_24h, rep(1:5, each = 4)) %>% as.data.frame()

#we combine the dataframe and order the groups
combined_data <- cbind(groups_6h, groups_24h) 
rownames(combined_data) <- c("rep 1", "rep 2", "rep 3", "rep 4")
colnames(combined_data) <- c("Unstimulated_6h", "GPVI_6h", "CLEC2_6h","TC_6h","NO_PEV_6h","Unstimulated_24h", "GPVI_24h", "CLEC2_24h","TC_24h", "NO_PEV_24h")

ordered_data <- combined_data[, c("CLEC2_6h", "CLEC2_24h", "GPVI_6h", "GPVI_24h", "TC_6h", "TC_24h", "Unstimulated_6h", "Unstimulated_24h", "NO_PEV_6h", "NO_PEV_24h")]

#we create a boxplot for total intensity both for 6h and 24h and color by group
par(mar = c(8,4,4,2) + 0.1)
boxplot(ordered_data, main = "Total intensities for PEV groups", names = c("CLEC2_6h", "CLEC2_24h", "GPVI_6h", "GPVI_24h", "TC_6h", "TC_24h", "Unstimulated_6h", "Unstimulated_24h", "No_PEV_6h", "No_PEV_24h"), las = 2, col = c("red","red" , "green","green" ,"cyan","cyan", "purple", "purple", "orange", "orange"))

```

## Performing MOFA on the cytokine data

```{r}

mofa_data <- list(view1 = cytokine_data)
lapply(mofa_data, dim)

#creating MOFA object
MOFA_object <- create_mofa_from_matrix(mofa_data)
views_names(MOFA_object) <- c("cytokines")
plot_data_overview(MOFA_object)
print(MOFA_object)

#Setting training options for data 
data_opts <- get_default_data_options(MOFA_object) 
head(data_opts)
#variances between views here do not differ drastically thus we do not need to scale views?

model_opts <- get_default_model_options(MOFA_object)
model_opts$num_factors <- 6
head(model_opts)

train_opts <- get_default_training_options(MOFA_object)
train_opts$convergence_mode <-"slow"
set.seed <- 42
head(train_opts)
#convergence mode fast is okay for exploration we would like to change it for the real training of the model to slow 

#We train the model 
MOFA_object <- prepare_mofa(
  object = MOFA_object,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

reticulate::use_condaenv("mofa_env10")
outfile = file.path("/home/aosterberg/Documents/DI_työ", "cytokinemodel.hdf5")
MOFAobject.trained <- run_mofa(MOFA_object, outfile=outfile, use_basilisk=FALSE)

```
We then load the model and extract the active factors into a csv file 
```{r}
model <- load_model(outfile, remove_inactive_factors = FALSE)
plot_variance_explained(model, x="view", y="factor")

factor1_df <- get_factors(model, factors = 1, as.data.frame = TRUE)
factor2_df <- get_factors(model, factors = 2, as.data.frame = TRUE)

factor1_df <- factor1_df %>% select(-group, -factor)
factor2_df <- factor2_df %>% select(-group, -factor)
head(factor1_df)

factors_df <- Reduce(function(x,y) merge(x,y, by = "sample"), list(factor1_df, factor2_df))
colnames(factors_df) <- c("sample", "Cytokine_Factor_1", "Cytokine_Factor_2")
head(factors_df)
write.csv(factors_df, "NEWcytokine_factors", row.names = FALSE)

```

