---
title: "PC and Factor contributions"
author: "Anna Österberg"
date: "2024-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## PC loadings for cytokines and weights for proteins in Factor 1 of top 200 MOFA model

```{r}
#Loading needed libraries
library(factoextra)
library(FactoMineR)
library(tidyverse)
library(MOFA2)
library(data.table)

```

We run a PCA on the cytokine data (24 h after treatment) including only PEV samples and extract contributions and loadings for PC1 and PC2 as csv files

```{r}
#We read in the data and assign samplenames as column names and cytokines as rownames and chose only PEV samples 
raw_data <- read.csv("/home/aosterberg/Documents/DI_työ/cytokine24h.csv",header = TRUE)
colnames(raw_data) <- c("secretome", "unstimulated_PEV_1", "unstimulated_PEV_2", "unstimulated_PEV_3", "unstimulated_PEV_4","GPVI_PEV_1", "GPVI_PEV_2", "GPVI_PEV_3", "GPVI_PEV_4", "CLEC2_PEV_1","CLEC2_PEV_2","CLEC2_PEV_3", "CLEC2_PEV_4", "TC_PEV_1","TC_PEV_2","TC_PEV_3", "TC_PEV_4", "no_PEV_1","no_PEV_2","no_PEV_3","no_PEV_4")
cytokine_data <- raw_data[,2:17] %>%  as.matrix()


#We run PCA
pca_data <- PCA(t(cytokine_data), scale.unit = TRUE, graph = FALSE)
summary(pca_data)

#We extract the contribution and loadings for each PC and order it in descending order based on contribution and save this as a csv file 
PC1contribution_df24h <- data.frame(contribution =  pca_data$var$contrib[,1], loadings = pca_data$var$coord[,1])[order(-pca_data$var$contrib[,1]),]
print(PC1contribution_df24h)
write.csv(PC1contribution_df24h, "PC1contribution.csv")

PC2contribution_df24h <- data.frame(contribution =  pca_data$var$contrib[,2], loadings = pca_data$var$coord[,2])[order(-pca_data$var$contrib[,2]),]
print(PC2contribution_df24h)
write.csv(PC2contribution_df24h, "PC2contribution.csv")

```

#PC loadings for cytokine data 6h after treatment


```{r}
# We extract 
cytokine_6h <- read_excel("/home/aosterberg/Documents/DI_työ/data/TableS2.xlsx", sheet = 2) %>% as.data.frame()
colnames(cytokine_6h) <- c("secretome", "unstimulated_PEV_1", "unstimulated_PEV_2", "unstimulated_PEV_3", "unstimulated_PEV_4","GPVI_PEV_1", "GPVI_PEV_2", "GPVI_PEV_3", "GPVI_PEV_4", "CLEC2_PEV_1","CLEC2_PEV_2","CLEC2_PEV_3", "CLEC2_PEV_4", "TC_PEV_1","TC_PEV_2","TC_PEV_3", "TC_PEV_4", "no_PEV_1","no_PEV_2","no_PEV_3","no_PEV_4")
rownames(cytokine_6h) <- cytokine_6h[,1]
PEVcytokine_6h <- cytokine_6h[,2:17] %>% as.matrix()

pca_data <- PCA(t(PEVcytokine_6h), scale.unit = TRUE, graph = FALSE)

PC1contribution_df6h <- data.frame(contribution =  pca_data$var$contrib[,1], loadings = pca_data$var$coord[,1])[order(-pca_data$var$contrib[,1]),]
print(PC1contribution_df6h)
write.csv(PC1contribution_df6h, "PC1contribution6h.csv")

PC2contribution_df6h <- data.frame(contribution =  pca_data$var$contrib[,2], loadings = pca_data$var$coord[,2])[order(-pca_data$var$contrib[,2]),]
print(PC2contribution_df6h)
write.csv(PC2contribution_df6h, "PC2contribution6h.csv")

```

Plotting the difference between 6h and 24h contributions as a barplot 

```{r}

#We create dataframes with the cytokines and the difference ordered by the difference between 24h and 6h contributions to PC1 and PC2
PC1_difference <-data.frame(Cytokine = rownames(PC1contribution_df24h), Difference = (PC1contribution_df24h$contribution - PC1contribution_df6h$contribution))[order(abs(PC1contribution_df24h$contribution - PC1contribution_df6h$contribution)),]
head(PC1_difference)

PC1_difference$Cytokine <- factor(PC1_difference$Cytokine, levels = PC1_difference$Cytokine) #we set the ordered cytokines as a factor so they are plotted in this order

#We repeat for PC2
PC2_difference <-data.frame(Cytokine = rownames(PC2contribution_df24h), Difference =(PC2contribution_df24h$contribution - PC2contribution_df6h$contribution))[order(abs(PC2contribution_df24h$contribution - PC2contribution_df6h$contribution)),]

PC2_difference$Cytokine <- factor(PC2_difference$Cytokine, levels = PC2_difference$Cytokine)


#We plot the PC difference on x-axis as barplots with the cytokines ordered from largest difference to smallest 
ggplot(PC1_difference, aes(x=PC1_difference$Difference, y = PC1_difference$Cytokine)) + geom_bar(stat="identity", fill = "purple") + theme_minimal() + labs(x= "Difference", y = "Cytokine", title= "Difference in PC1 contributions")+ theme(plot.title = element_text(hjust=0.5, size =20))

ggplot(PC1_difference, aes(x=PC2_difference$Difference, y = PC2_difference$Cytokine)) + geom_bar(stat="identity", fill = "purple") + theme_minimal() + labs(x= "Difference", y = "Cytokine", title= "Difference in PC2 contributions")+ theme(plot.title = element_text(hjust=0.5, size =20))

```


Since we saw that the top 200 model Factor 1 vs PC2 showed most promise for the PEV samples clustering togheter by groups and in a linear fashion we thus only study the factor contributions for the first factor of that model

```{r}
model200 <- load_model("/home/aosterberg/Documents/DI_työ/top200MOFAmodel.hdf5") #we load the model


#We plot the weights for both proteins and miRNA
plot_weights(model200, view = "proteins", scale= FALSE, factor = 1) 
plot_weights(model200, view = "miRNA", factor = 1, scale = FALSE) 

#we extract the weights from each view as a dataframe and order them by absolute value in decreasing order
prot_weights <- get_weights(model200, view = "proteins", factor = 1, scale = TRUE, as.data.frame = TRUE)
miRNA_weights <- get_weights(model200, view = "miRNA", factor = 1, scale= TRUE, as.data.frame = TRUE)

ordered_miRNA <- miRNA_weights[order(-abs(miRNA_weights$value)),] %>% select(-view, -factor)
head(ordered_miRNA)
ordered_prot <- prot_weights[order(-abs(prot_weights$value)),] %>% select(-view, -factor)
head(ordered_prot)

# we filter the miRNAs with a weight above 0.5 and see that there are none therefore we will from now on study proteins
filtered_miRNA <- ordered_miRNA %>% filter(abs(value)>0.5)
print(filtered_miRNA)


#we create a csv file with all proteins with a weight above 0.5 as well as one for all proteins with a positive weight above 0.5 and a weight less than -0.5
filtered_prot <- ordered_prot %>% filter(abs(value)>0.5)
neg_vals_0.5 <- filtered_prot %>% filter(value < 0)
pos_vals_0.5 <- filtered_prot %>% filter(value > 0)

write.csv(neg_vals_0.5, "neg_vals_0.5.csv", row.names = FALSE)
write.csv(pos_vals_0.5, "pos_vals_0.5.csv", row.names = FALSE)
write.csv(filtered_prot, "top_proteins_0.5.csv", row.names = FALSE)

```


