---
title: "PEA_analysis"
author: "Anna Österberg"
date: "2024-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE)
```

## PEA (inflammation panel) analysis 
-creating factors 
-running PCA, extracting PC loading s
-finding top 10 proteins contributing to variance between groups 
-plotting these as both barplots and boxplots colored by sample groups 

```{r}
#Loading needed libraries 
library(ggplot2)
library(factoextra)
library(FactoMineR)
library(tidyverse)
library(MOFA2)
library(reticulate)
library(data.table)

#Loading and formatting the data
raw_data <- read.csv("/home/aosterberg/Documents/DI_työ/PEVinflammationpanel.csv",header = TRUE)
rownames(raw_data) <- raw_data[,1]
colnames(raw_data) <- c("protein", "US_PEV_1","US_PEV_2","US_PEV_3", "US_PEV_4", "GPVI_PEV_1", "GPVI_PEV_2", "GPVI_PEV_3", "GPVI_PEV_4", "CLEC2_PEV_1","CLEC2_PEV_2","CLEC2_PEV_3","TC_PEV_1","TC_PEV_2","TC_PEV_3")
PEA_df <- raw_data[,2:15] 
PEA_data <- as.matrix(PEA_df)

```
### Normalizing the data

```{r}
#We create boxplots for both unscaled and scaled data
boxplot(PEA_data[,-1], cex.axis = 1.2, cex.names = 1) 
boxplot(PEA_data[,-1] %>% scale, cex.axis = 1.2, cex.names = 1)


#We create a function to normalize to the standard distribution 
stdnorm <- function(x){
  r = rank(x, ties.method = "random")
  qnorm(r / (length(x) +1))
}

#We create matrices of the data and apply the function above
PEA_norm <- apply(PEA_data, 2, stdnorm)
boxplot(PEA_norm, cex.axis = 1.2)

#Adding groups for coloring the samples in the plot 
PEA_groups <- c("US", "US","US", "US", "GPVI", "GPVI", "GPVI", "GPVI", "CLEC2", "CLEC2", "CLEC2", "TC", "TC","TC")
sample_colors <- c("CLEC2" = "red", "GPVI" = "green", "TC" = "blue", "US"= "purple")


pca_data <- PCA(t(PEA_norm), scale.unit = TRUE, graph = FALSE)
fviz_eig(pca_data, addlabels = TRUE)
eigenvalues <- pca_data$eig 

#Plotting the PCA data, coloring by groups and showing variance explained by each principal component
fviz_pca_ind(pca_data,
             col.ind = PEA_groups, # color by groups
             palette = sample_colors, mean.point = FALSE, repel = TRUE,
             legend.title = "Groups")+labs(title = "PCA of  normalized PEA data", x= paste0("PC1 (",round(eigenvalues[1,2],2),"%)"), y= paste0("PC2 (",round(eigenvalues[2,2],2),"%)")) + theme_minimal()+ scale_shape_manual(values = rep(16, length(unique(PEA_groups)))) + guides(color = guide_legend(override.aes = list(shape = 16, size = 4)))+theme(axis.title.x = element_text(size = 13, color = "black"), axis.title.y = element_text(size = 13, color = "black"), axis.text.x = element_text(size = 13, color = "black"), axis.text.y = element_text(size = 13,  color = "black"),legend.text = element_text(size = 12), legend.title = element_text(size  = 12))
                                                                                                                                                                                                                                                                                                                                                       

#creating a list of cytokines and plotting each and saving in a pdf
cytokines_6h <- unique(cytokines_long_6h$Cytokine)


```

The normalization still keeps the separation between groups compared to normalizing cytokine data with same steps, thus we move forward with normalized dataset and extract the loadings and factor weights using this data. 


### Running PCA on PEA data 
  -extracting loadings 

```{r}

pca_data <- PCA(t(PEA_norm), scale.unit = TRUE, graph = FALSE)
#summary(pca_data)
fviz_eig(pca_data, addlabels = TRUE) 

#Since PC1 explains 51.9% of the variance and PC2 16.6% we extract the loadings for these two 
PC1loadings <- data.frame(loading = pca_data$var$coord[,1]) %>% arrange(-abs(loading))
print(PC1loadings)
PC2loadings <- data.frame(loading = pca_data$var$coord[,2]) %>% arrange(-abs(loading))
print(PC2loadings)
```

### Running MOFA on PEA data and extracting factors and weights 

```{r}

mofa_data <- list(view1 = PEA_norm)
lapply(mofa_data, dim)

#creating MOFA object
MOFA_object <- create_mofa_from_matrix(mofa_data)
views_names(MOFA_object) <- c("inflammation panel")
plot_data_overview(MOFA_object)
print(MOFA_object)

#Setting training options for data 
data_opts <- get_default_data_options(MOFA_object) 
head(data_opts)


model_opts <- get_default_model_options(MOFA_object)
model_opts$num_factors <- 6
head(model_opts)

train_opts <- get_default_training_options(MOFA_object)
train_opts$convergence_mode <-"slow"
set.seed <- 42
head(train_opts)


#We train the model 
MOFA_object <- prepare_mofa(
  object = MOFA_object,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

#In terminal conda create -n mofa_env python=3.10, conda activate mofa_env (you don't have to create a new environment each time, reticulate package you have downloaded does not seem to work with python 3.12 use python 3.10)
reticulate::use_condaenv("mofa_env10")
outfile = file.path("/home/aosterberg/Documents/DI_työ","PEAmodel.hdf5")
MOFAobject.trained <- run_mofa(MOFA_object, outfile=outfile, use_basilisk=FALSE)


##############################################################################################################################
#Downstream analysis of the trained model 
model <- load_model(outfile, remove_inactive_factors = TRUE)
plot_variance_explained(model, x="view", y="factor")

```

### Loading PEA data for top proteins and plotting boxplots 

```{r}
PEA_weights_factor1 <- get_weights(model, view = "inflammation panel", factor = 1, scale = FALSE, as.data.frame = TRUE)  %>%  arrange(-abs(value))
print(PEA_weights_factor1)

```

### Boxplots of the top proteins from loadings in PC1 and PC2 as well as weights for factor 1 

```{r}

#Creating dataframes for the overrepresented proteins to be plotted
PC1data <- PEA_df[rownames(PC1loadings)[1:10],]
PC2data <- PEA_df[rownames(PC2loadings)[1:10],]
Factor1data <-PEA_df[PEA_weights_factor1$feature[1:10],]
Factor1data


#Formatting the dataframes for plotting
PC1_long <- PC1data %>% 
  rownames_to_column(var = "Protein") %>% 
  pivot_longer(-Protein, names_to= "Sample", values_to = "NPX")%>% 
  mutate(Group=rep(PEA_groups, times  = 10))

PC2_long <- PC2data %>% 
  rownames_to_column(var = "Protein") %>% 
  pivot_longer(-Protein, names_to= "Sample", values_to = "NPX")%>% 
  mutate(Group=rep(PEA_groups, times  = 10))

Factor1_long <- Factor1data %>% 
  rownames_to_column(var = "Protein") %>% 
  pivot_longer(-Protein, names_to= "Sample", values_to = "NPX")%>% 
  mutate(Group=rep(PEA_groups, times  = 10))



#Function for barplot of each protein colored by group
boxplot_proteins <- function(df_long, name){
  df_long %>% filter(Protein == name) %>% 
  ggplot(aes(x= Group, y = NPX,  fill = Group))+ geom_boxplot() + theme_minimal()+ labs(title = name)+guides(color = guide_legend(override.aes = list(size = 4)))+theme(axis.title.x = element_text(size = 13, color = "black"), axis.title.y = element_text(size = 13, color = "black"), axis.text.x = element_text(size = 13, color = "black"), axis.text.y = element_text(size = 13,  color = "black"),legend.text = element_text(size = 12), legend.title = element_text(size  = 12), plot.title = element_text(size = 15, hjust = 0.5))
}

#Creating lists of proteins to be plotted and plotting them in a pdf

PC1_proteins <- unique(PC1_long$Protein)
PC2_proteins <- unique(PC2_long$Protein)
factor1_proteins <- unique(Factor1_long$Protein)


pdf("PC1_boxplots.pdf")
for(protein in PC1_proteins){
  print(boxplot_proteins(PC1_long,protein))
}
dev.off()

pdf("PC2_boxplots.pdf")
for(protein in PC2_proteins){
  print(boxplot_proteins(PC2_long,protein))
}
dev.off()


pdf("Factor1_boxplots.pdf")
for(protein in factor1_proteins){
  print(boxplot_proteins(Factor1_long,protein))
}
dev.off()
```


```


