---
title: "Comparing PCs and Factors"
author: "Anna Österberg"
date: "2024-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

##Plotting PCs against factors 

```{r}
#Loading needed libraries 
library(factoextra)
library(FactoMineR)
library(ggplot2)

#We perform PCA on the macrophage secretome of differently activated PEVs
raw_data <- read.csv("/home/aosterberg/Documents/DI_työ/cytokine24h.csv",header = TRUE)
colnames(raw_data) <- c("secretome", "unstimulated_PEV_1", "unstimulated_PEV_2", "unstimulated_PEV_3", "unstimulated_PEV_4","GPVI_PEV_1", "GPVI_PEV_2", "GPVI_PEV_3", "GPVI_PEV_4", "CLEC2_PEV_1","CLEC2_PEV_2","CLEC2_PEV_3", "CLEC2_PEV_4", "TC_PEV_1","TC_PEV_2","TC_PEV_3", "TC_PEV_4", "no_PEV_1","no_PEV_2","no_PEV_3","no_PEV_4")
test_data <- raw_data[,2:17]
cytokine_data <- as.matrix(test_data)
pca_data <- PCA(t(cytokine_data), scale.unit = TRUE, graph = FALSE)
summary(pca_data)

#We choose the 3 first PCs and create a dataframe with their values
fviz_eig(pca_data, addlabels = TRUE) 
head(pca_data)
pca_components <- data.frame(PC1 =pca_data$ind$coord[,1], PC2 =pca_data$ind$coord[,2],PC3 =pca_data$ind$coord[,3])
head(pca_components)


#we load the dataframe containing the factors and sort them so samples are in the same orders in both dataframes
factors_df <- read.csv(file.choose())
rownames(factors_df)<- factors_df[,1]
order <- c("unstimulated_PEV_1", "unstimulated_PEV_2", "unstimulated_PEV_3", "unstimulated_PEV_4","GPVI_PEV_1", "GPVI_PEV_2", "GPVI_PEV_3", "GPVI_PEV_4", "CLEC2_PEV_1","CLEC2_PEV_2","CLEC2_PEV_3", "CLEC2_PEV_4", "TC_PEV_1","TC_PEV_2","TC_PEV_3", "TC_PEV_4")
sorted_factors <- factors_df[match(order, rownames(factors_df)),]
head(sorted_factors)


#We create a combined dataframe and format it into matrix for plotting 
factors_pca <- cbind(pca_components,sorted_factors[,-1])
head(factors_pca)
factor_pca_mat <- as.matrix(factors_pca)
pairs(factor_pca_mat, main = "Top 200 protein and miRNA Factors vs cytokine PCs")

```

## We also plot MOFA factors for cytokines against the protein and miRNA factors 
```{r}
#we load the dataframe containing the protein and miRNA factors and order it
miRNAprotfactors_df <- read.csv(file.choose())
rownames(miRNAprotfactors_df)<- miRNAprotfactors_df[,1]
order <- c("unstimulated_PEV_1", "unstimulated_PEV_2", "unstimulated_PEV_3", "unstimulated_PEV_4","GPVI_PEV_1", "GPVI_PEV_2", "GPVI_PEV_3", "GPVI_PEV_4", "CLEC2_PEV_1","CLEC2_PEV_2","CLEC2_PEV_3", "CLEC2_PEV_4", "TC_PEV_1","TC_PEV_2","TC_PEV_3", "TC_PEV_4")
miRNAprotsorted_factors <-miRNAprotfactors_df[match(order, rownames(miRNAprotfactors_df)),]
head(miRNAprotsorted_factors)

#We load the file containing cytokine factors and order it 
cytokinefactors_df <- read.csv(file.choose())
rownames(cytokinefactors_df) <- cytokinefactors_df[,1]
cytokine_sorted_factors <-cytokinefactors_df[match(order, rownames(cytokinefactors_df)),]
head(cytokine_sorted_factors)

#we create a combined dataframe and convert it to a matrix
combined_factors <- cbind(cytokine_sorted_factors[,-1],miRNAprotsorted_factors[,-1])
head(combined_factors)
combined_mat <- as.matrix(combined_factors)

#we plot the factors against each other
pairs(combined_mat)
```

## Plotting a single factor against PC
-based on the plots above we choose single factors (prot and miRNA) to plot against a PC(cytokine)
-we then fit a linear regression to the data and print the p-value 
-we also try fitting a linear regression to the sum of all factors (prot and miRNA) against a PC(cytokine)

```{r}

#Plotting single factors and PCs against each other
groups <- c("unstimulated", "unstimulated", "unstimulated", "unstimulated", "GPVI", "GPVI","GPVI", "GPVI", "CLEC2", "CLEC2", "CLEC2", "CLEC2","TC","TC","TC","TC")
factorPC_df <- as.data.frame(factor_pca_mat)
factorPC_df$group <- groups
ggplot(factorPC_df, aes(x=Factor4, y=PC3, color = group)) +geom_point(size  = 2.5 )+ theme_minimal()


#Fitting a linear regression model to the PC vs Factor 
modelFA1_PC2 <- lm(PC2 ~ Factor1 , data = factorPC_df)
summary(modelFA1_PC2)
ggplot(factorPC_df, aes(x=Factor1, y=PC2, color = group)) +geom_point(size  = 2.5 ) + geom_smooth(method = "lm", color = "black")+ theme_minimal()

modelFA4_PC3 <- lm(PC3 ~ Factor4 , data = factorPC_df)
summary(modelFA4_PC3)
ggplot(factorPC_df, aes(x=Factor4, y=PC3, color = group)) +geom_point(size  = 2.5 ) + geom_smooth(method = "lm", color = "black")+ theme_minimal()


#Fitting a regression model to the sum of the factors 
modelPC2FA <- lm(PC2 ~ (Factor1 + Factor2 + Factor3 + Factor4 + Factor5 + Factor6), data = factorPC_df)
summary(modelPC2FA)
modelPC3FA <- lm(PC3 ~ (Factor1 + Factor2 + Factor3 + Factor4 + Factor5 + Factor6), data = factorPC_df)
summary(modelPC3FA)

       
```






